How explicit is the Explicit Formula?

### 2016-08-02 paper/paper.tex
paper: revised Aug 1 version

%\hskip20pt ({\it Notes for our  20+20 minute talk at the  AMS Special Session on Arithmetic Statistics in San Diego, January 2013}
\vskip10pt
\tableofcontents
\section{Explicit Formulas}
\vskip10pt

@@

In celebration of Riemann's formula  we will discuss some open questions related to the application of the Explicit Formula to the arithmetic of elliptic curves. In this context, a cartoon version of the genre of `Explicit Formula' might be described as follows:

 {{\bf (**)} \ \  ${\rm{\it Sum\ of\ local\ data\  }}  \ = \  {\rm{\it Global\ datum}}\ +\  {\rm{\it  Easy\ continuous\ error\ term}}\ +\  {\rm{\it  Oscillatory\ term}},$ }

@@

[[For example, for 11a or 37a or 389a ??  and show a plot that anybody could
reproduce, which raises questions.]]

   Usually the {\it Global datum} is computed by knowing the order of specific zeroes (or poles) at specified values of the complex variable $s$  of relevant (global) $L$-functions.  For example, the formula {(*)} gives the contribution of the pole at $s=1$ of $\zeta(s)$.  Often (perhaps only conjecturally)  the  {\it Global datum} is the `dominant term'    on the right-hand-side of the equation.  Sometimes, as will be the case  of the examples to be described below, the roles are reversed and one takes the Global datum as the object being sought.  That is, we view it as ${\rm{\it Global\ datum}}\ = \  {\rm{\it Sum\ of\ local\ data\  }}\ -\  {\rm{\it  Easy\ error\ term}}\ -\  {\rm{\it  Oscillatory\ term}}.$

   In some instances,  the sought-for Global datum is  (conjecturally) constant, independent of the cutoff $X$---in fact, the mean of {\it Sum of local data}---and also an integer, in which case a close approximation to each of the other three terms (for some specific value of $X$) would give a decisive answer for the value of the Global datum.   This has been a theme in some recent work, especially in the context of elliptic curves over ${\Q}$.
   For this volume in honor of Bernhard Riemann,  we will examine one interesting case  that does depend on these `finer conjectures',  advertise the conjectures themselves, and  the need for  computational projects regarding  such problems.  We restrict ourselves to applications to elliptic curves, but we take this only as a basic (important) example of a fuller story for general motives.  We have  future plans for a web-accessible resource\footnote{See \url{http://wstein.org/papers/2016-explicit/}}: a repository of some of the numerics for the cases related to elliptic curves that interest us.

  Specifically we will  focus on {\it issues of bias}  following the classical ``Explicit Formula,'' and the work of:
Sarnak, \ Granville,\  Rubenstein,\, Martin-Watkins,\ Fiorilli,\ Bober-Mestre-Odlyzko,\ Conrey-Snaith, \ Spicer, and others.

 The {\it example problems} we consider are related to the question---given an elliptic curve over the rational numbers and letting  $p$ range through prime numbers---of how often  $p+1$ is an {\it over-count} or an {\it under-count} for the number of rational points on the curve modulo $p$? The rough answer is 50/50, but there can be a `bias' similar to the classical Chebyshev bias  \cite{R-S}. For such finer statistics one resorts to `Explicit Formulas.'  Here, computation can  outstrip theory in that people have algorithms to make precise computations (their correctness being conditional on some conjectures, including the holomorphicity of the $L$-functions in question).

 We offer no new theoretical results but use this occasion to mention some interesting recent work and conjectures  (of other people)  that might warrant more such computations and that raise a host of questions, both theoretical and computational.   For example, to do some systematic numerical computations related to an elliptic curve $E$ attached to a newform $f_E$  (along the lines of what has already been done in this paper)  it would be very useful to have a much larger data set  of the arithmetic function  $$n \mapsto r_E(n)$$
 where $r_E(n)$ is the order of vanishing [[where? at what point?]] of the $L$-function of the automorphic forms ${\rm symm}^n f_E$ for odd values of $n$.  Regarding this arithmetic function, aside from having control of the parity of   $r_E(n)$  (e.g., see \cite{DMW})  hardly anything else is known. Nor do we (at least, the authors of this paper)  yet have enough
 experience---when $E$ has no complex multiplication---even to formulate a proper conjecture.

 We might also mention that when making these numerical experiments one seems to be in a situation  that is not entirely dissimilar from the type of slightly annoying mismatch between conjecture and data that one encounters in more traditional studies of Mordell-Weil statistics  that was the subject of the survey article \cite{bmsw:bulletins}, and the more recent \cite{ecdb:height}.  But this may be unavoidable, given that even  the so-called  `easy error term' in the explicit formula will tend to converge  only $O(1/\log X)$ fast.

@@

\subsection{ An example}   To give an example, consider a non-CM elliptic curve $E$ over ${\Q}$ of conductor $N$  uniformized by a newform $f_E$ (of conductor $N$ and of weight two) with Fourier expansion  $f_E(q) = \sum_{n\ge 1}a_n(E)q^n$. Form the following {\it Sum of local data}:

    $${D}_E(X):= {\frac{1}{\log\ X}}\sum_{p \le X}{\frac{a_p(E)\log p}{ p}}$$

    which fits into this Explicit Formula (see \cite[Formula (vi) of the appendix]{S}).{\footnote{ There is a sign typo there: $r_E$ should be $-r_E$.}}) dependent on GRH together with some other conjectures
(for these, see *** below).

    $${\bf (*)}\ \ \ {D}_E(X) = -r_E \ \ \  +\ \ \  c(E)/\log\ X  +o(1/\log X)   \ \ \  +\ \ \
     {\frac{1}{\log X}}\lim_{T\to \infty}S_E(X;T).$$

     Here,

     \begin{itemize} \item   $r_E$ is the analytic rank of $E$  (i.e., the order of vanishing of  $L(E,s)$ at the central point; conjecturally this will be the Mordell-Weil rank of $E$ )\item $c(E)$ is a constant, and \item the last term  on the right is the `Oscillatory term' where  $S_E(X;T):=\sum_{|\gamma | \le T}X^{i\gamma}/i\gamma $, the sum over $\gamma$ ranging through the nonzero imaginary parts of the zeroes of $L(E,s) $.\end{itemize}

     Subject to the further conjecture  (see \cite{S})  that the Oscillatory term goes to zero as $X$ goes to infinity, i.e., $$ {\frac{1}{\log X}} \lim_{T\to \infty}\sum_{|\gamma | \le T}X^{i\gamma}/i\gamma  \ = \ o(1),$$  one obtains that  ${D}_E(X)$ achieves a limiting distribution of values [[forward reference to section below; actually compute and discuss this distribution for ${D}_E(X)$ as well]], with mean equal to $-r_E$.  It is natural to view this in the context of Chebyshev biases; i.e., to interpret---when $r_E$ is positive---the minus sign in $-r_E$ as a measure of the bias that the values $a_p(E)$ have to being negative. This type of {\it bias} in the arithmetic statistics of elliptic curves  hearkens back to the early work of Birch and Swinnerton-Dyer \cite{S}, and in the context we are considering was  first written down by Peter Sarnak  in the spirit of the classical Chebyschev bias, and 'prime races;' an `Explicit Formula' account of this in the context of  prime numbers in different arithmetic progressions can be found in {\cite{GM}}.

@@

    One might ask for more explicit forms of the above equation {\bf (*)}.   For example, consider the following basic question.  For a real number $x$,  let  $\|x\|$ denote the distance between $x$ and the nearest integer. [[think of a better notation for this  -- too much like norm]]

   \begin{quote}  Find a function $N \mapsto X(N)$ such that for all $X \ge X(N)$, and for any elliptic curve $E$ over ${\Q}$ of conductor $N$ one has   the inequality
   $$\|{\frac{1}{\log\ X}}\sum_{p \le X}{\frac{a_E(p)\log p}{ p}}\| \ \ < \ \ 1/2.$$ \end{quote}

 That there is such a function $N \mapsto X(N)$ is already conjectural. Its existence would offer a simple formula for the analytic rank of $E$  in terms of the $a_p(E)$ for $p < X(N)$.   That is, $-r_E$ would be the integer nearest any of the values $D_E(X)$ for $X >X(N)$.   For a detailed study of computer complexity issues related to the computation of   $a_p(E)$ and $r_E$, see (Sarnak and ***).

@@

    \subsection{The Oscillatory term}\label{osc}
   The analogous oscillatory term for the classical Riemann zeta function  has an extensive literature. See, for example \cite{G}, \cite{Fu} and the bibliography there. Here are some pictures to convey a sense of how our $S_E(X;T)$ behaves, at least in the currently computable range which (roughly) allows $T$ to be only as high as  $10^4$.

  \centerline{ $E = 11a$}
 \vskip10pt
 \includegraphics[width=0.9\textwidth]{plots/OSC11.pdf}

$$U_n(\theta) : = {\frac {\sin(n+1)\theta}{\sin\theta}}$$ and note that the set $\{U_n\}$ for $n=0,1,2,\dots$ forms an orthonormal basis of the Hilbert space $L^2[0,\pi]$.

\begin{definition} Say that our data (\ref{data}) has {\bf `Explicit Formula' statistics} if there is a sequence of non-negative integers $\{r_n\}_n$  for $n=1,2,3, \dots$ such that for all smooth functions $V(\theta)$ as above with $c_0=0$, the ``$V$-weighted average of the data"
\begin{equation}
S_V(X):= {\frac{\log X}{\sqrt X}}\sum_{p \le X} \ V(\theta_p)
\end{equation}
\begin{itemize}
\item
possesses a limiting distribution{\footnote{\label{footnote:statdist} Recall that, as in subsection    \ref{statdist} above,  $S_V(x)$ {\bf possesses a limiting distribution $\mu_V$ with respect to the multiplicative measure $dx/x$} if for continuous bounded functions $f$ on ${\bf R}$ we have:
\begin{equation}
\lim_{X \to {\infty}}\ {\frac{1}{\log X}}\int_0^Xf(S_V(x))dx/x \ = \ \int_{\bf R}f(x)d\mu_V(x).
\end{equation}}}
 $\mu_V$ with respect to the multiplicative measure $dX/X$,
\item  $\mu_V$ has support on all of ${\bf R}$, is continuous and symmetric about its mean, ${\mathcal E}(S_V)$, and
\begin{equation}\label{eqnmean}
{\mathcal E}(S_V)\ = \ -\sum_{n=1}^{\infty}  c_n\big(2r_n+(-1)^n\big).
\end{equation}
\end{itemize}
\end{definition}

\bigskip
One can also compute---given some plausible conjectures---the behavior of the {\bf variance}  (i.e., the measure of fluctuation of the values of $S_V(X)$ about the mean) as well; the variance is defined by the formula  $${\mathcal V}(S_V): = {\mathcal E}\big([S_V  - {\mathcal E}(S_V)]^2\big).$$

\subsection{The bias between under-counts and over-counts}
  We will assume that our data has `Explicit Formula' statistics, and---copying Sarnak ({\cite{S}})--- apply this to the question we began with, i.e., what is the ``bias" in the race between under-counts and over-counts?

@@

[[idea -- instead just define $\Phi(H,T)$ to be the min
of the finitely many possible linear combinations, and ask
about the behavior of $\Phi$.  Is there some clever algorithm
to get information about this?]]
\end{itemize}\end{itemize}
 %More specifically, the {\it means}  directly interpretable as {\it biases}  determine   the orders of vanishing at the central point of the $L$-functions of the symmetric powers of odd order of the modular eigenform attached to $E$ and vice versa..

>>>>

%\hskip20pt ({\it Notes for our  20+20 minute talk at the  AMS Special Session on Arithmetic Statistics in San Diego, January 2013}
\vskip10pt
%\tableofcontents
\section{Introduction}
\vskip10pt

@@

In celebration of Riemann's formula  we will discuss some computations and open questions related to the application of the Explicit Formula to the arithmetic of elliptic curves. This follows constructions in a letter of Peter Sarnak to one of the authors (\cite{S}).

A cartoon version of the genre of `Explicit Formula' might be described as follows:

 {{\bf (**)} \ \  ${\rm{\it Sum\ of\ local\ data\  }}  \ = \  {\rm{\it Global\ datum}}\ +\  {\rm{\it  Easy\  error\ term}}\ +\  {\rm{\it  Oscillatory\ term}},$ }

@@

   Usually the {\it Global datum} is computed by knowing the order of specific zeroes (or poles) at specified values of the complex variable $s$  of relevant (global) $L$-functions.  For example, the formula {(*)} gives the contribution of the pole at $s=1$ of $\zeta(s)$.  Often (perhaps only conjecturally)  the  {\it Global datum} is the `dominant term'    on the right-hand-side of the equation.  Sometimes, as will be the case  of the examples to be described below, the roles are reversed and one takes the Global datum as the object being studied.  That is, we view it as ${\rm{\it Global\ datum}}\ = \  {\rm{\it Sum\ of\ local\ data\  }}\ -\  {\rm{\it  Easy\ error\ term}}\ -\  {\rm{\it  Oscillatory\ term}}.$

   In some instances,  the sought-for Global datum is  (conjecturally) constant, independent of the cutoff $X$---in fact, the mean of {\it Sum of local data}---and also an integer, in which case a close approximation to each of the other three terms (for some specific value of $X$) would give a decisive answer for the value of the Global datum.

 \subsection{\bf An example: }  Consider a non-CM elliptic curve $E$ over ${\Q}$ of conductor $N$  uniformized by a newform $f_E$ (of conductor $N$ and of weight two) with Fourier expansion  $f_E(q) = \sum_{n\ge 1}a_n(E)q^n$. Form the following {\it Sum of local data}:

    $$ {\mathcal D}_E(X): = {\frac{\log\ X}{\sqrt X}}\sum_{p \le X}{\frac{ a_{E}(p)}{\sqrt p}},$$

    \noindent where (assuming the Riemann Hypothesis for the $L$-function attached to  $f_E$) the explicit formula gives us:

    $$ {\mathcal D}_E(X)  \ = \ 1-2r_E  \ +\  {\rm{\it  Easy\  error\ term}}\ +\  \sum_{|\gamma| \le T}{\frac{X^{i\gamma}}{{\frac{1}{2}}+i\gamma}}$$

     Here, the `global datum'  is $1-2r_E$ where  $r_E$ is the analytic rank of $E$  (i.e., the order of vanishing of  $L(E,s)$ at the central point; conjecturally this will be the Mordell-Weil rank of $E$). The summation is taken for $\gamma$ ranging over the imaginary parts of the nontrivial zeroes of the $L$-function attached to $E$.

   The   values of  ${\mathcal D}_E(X)$ achieve a  limiting distribution $\mu_E$ (with respect to multiplicative Haar measure $dx/x$) with {\it mean} $M$ equal to $1- 2r_E$  and variation  $V$  equal to $$\lim_{T \to \infty} \sum_{|\gamma | \le T}{\frac{1}{{\frac{1}{4}} + \gamma^2}}.$$

   Under further conjectures (\cite{S})  the measure $\mu_E$  is symmetric
about its mean $1-2r_E$, it is smooth, and has support all of $(-\infty, +\infty)$. It is natural to view this in the context of Chebyshev biases; i.e., to interpret---when $r_E$ is positive---the minus sign in $-r_E$ as a measure of the bias that the values $a_p(E)$ have to being negative. This type of {\it bias} in the arithmetic statistics of elliptic curves  hearkens back to the early work of Birch and Swinnerton-Dyer \cite{S}, and in the context we are considering was  first written down by Peter Sarnak  in the spirit of the classical Chebyschev bias, and 'prime races.'   An `Explicit Formula' account of this in the context of  prime numbers in different arithmetic progressions can be found in {\cite{GM}}.  See also \cite{R-S}, \cite{S}
which fits into this Explicit Formula format, as does  work of  Martin-Watkins,\ Fiorilli,\ Bober-Mestre-Odlyzko,\ Conrey-Snaith, \ Spicer, and others.

   For this volume in honor of Bernhard Riemann,  we will offer computations related to this, and one  other interesting case  that does depend on these `finer conjectures',  advertise the conjectures themselves, and  the need for  computational projects regarding  such problems.  We restrict ourselves to applications to elliptic curves, but we take this only as a basic (important) example of a fuller story for general motives.  We have  future plans for a web-accessible resource\footnote{See \url{http://wstein.org/papers/2016-explicit/}}: a repository of some of the numerics for the cases related to elliptic curves that interest us.

 One problem we consider is related to the question---given an elliptic curve over the rational numbers and letting  $p$ range through prime numbers---of how often  $p+1$ is an {\it over-count} or an {\it under-count} for the number of rational points on the curve modulo $p$? The rough answer is 50/50, but there can be a `bias'.

 We offer no new theoretical results but, as mentioned, we use this occasion to exhibit computations and recall some interesting recent work and conjectures  (of other people)  that might warrant more such computations and that raise a host of questions, both theoretical and computational.   For example, to do some systematic numerical computations related to an elliptic curve $E$ attached to a newform $f_E$  (along the lines of what has already been done in this paper)  it would be very useful to have a much larger data set  of the arithmetic function  $$n \mapsto r_E(n)$$
 where $r_E(n)$ is the order of vanishing at $s=1$ of the $L$-function of the automorphic forms ${\rm symm}^n f_E$ for odd values of $n$.  Regarding this arithmetic function, aside from having control of the parity of   $r_E(n)$  (e.g., see \cite{DMW})  hardly anything else is known. Nor do we (at least, the authors of this paper)  yet have enough
 experience---when $E$ has no complex multiplication---even to formulate a proper conjecture.

 We might also mention that when making these numerical experiments one seems to be in a situation  that is not entirely dissimilar from the type of slightly annoying mismatch between conjecture and data that one encounters in more traditional studies of Mordell-Weil statistics  that was the subject of the survey article \cite{bmsw:bulletins}, and the more recent \cite{ecdb:height}.  But this may be unavoidable, given that even  the so-called  `easy error term' in the explicit formula may tend to zero rather slowly.

@@

 \section{\bf Sarnak distributions for the elliptic curve $E$ relative to the weighting function $V$}

  In a letter  \cite{S}  to one of us (to B.M.) Peter Sarnak considered a broad array of statistics related to local arithmetic data associated to modular forms,  and, in particular, to elliptic curves over ${\Q}$.

$$U_n(\theta) : = {\frac {\sin(n+1)\theta}{\sin\theta}}$$  so we have:

  $$\frac{a_E(p)}{\sqrt p} = U_1(\theta_{E,p}).$$  Note that the set $\{U_n\}$ for $n=0,1,2,\dots$ forms an orthonormal basis of the Hilbert space $L^2[0,\pi]$  with the
 inner product

 $$\langle f, g\rangle :={\frac{2}{\pi}} \int_0^\pi f(\theta)g(\theta)\sin^2(\theta)d\theta.$$

Let us define the ``$V$-weighted average of the data"
\begin{equation}
D_V(X):= {\frac{\log X}{\sqrt X}}\sum_{p \le X} \ V(\theta_p)
\end{equation}

 so that   $${\mathcal D}_E(X)=  {\mathcal D}_{U_1}(X).$$

\begin{definition} Say that our data (\ref{data}) has {\bf `Explicit Formula' statistics} if there is a sequence of non-negative integers $\{r_n\}_n$  for $n=1,2,3, \dots$ such that for all smooth functions $V(\theta)$ as above with $c_0=0$,
\begin{itemize}
\item
possesses a limiting distribution{\footnote{\label{footnote:statdist} Recall that, as in subsection    \ref{statdist} above,  $S_V(x)$ {\bf possesses a limiting distribution $\mu_V$ with respect to the multiplicative measure $dx/x$} if for continuous bounded functions $f$ on ${\bf R}$ we have:
\begin{equation}
\lim_{X \to {\infty}}\ {\frac{1}{\log X}}\int_0^Xf(S_V(x))dx/x \ = \ \int_{\bf R}f(x)d\mu_V(x).
\end{equation}}}
 $\mu_V$ with respect to the multiplicative measure $dX/X$,
\item  $\mu_V$ has support on all of ${\bf R}$, is continuous and symmetric about its mean, ${\mathcal E}(S_V)$, and
\begin{equation}\label{eqnmean}
{\mathcal E}(S_V)\ = \ -\sum_{n=1}^{\infty}  c_n\big(2r_n+(-1)^n\big).
\end{equation}
\end{itemize}
\end{definition}
We will refer to $\mu_V$  as the {\bf Sarnak Distribution  (for the elliptic curve $E$ relative to the weighting function $V$)}.
One can also compute---given some plausible conjectures---the behavior of the {\bf variance}  (i.e., the measure of fluctuation of the values of $S_V(X)$ about the mean) as well; the variance is defined by the formula  $${\mathcal V}(S_V): = {\mathcal E}\big([S_V  - {\mathcal E}(S_V)]^2\big).$$

@@

 {\bf The bias of undercounts versus overcounts  (\cite{S})}:

@@

  %  \subsection{The Oscillatory term}\label{osc}
  % The analogous oscillatory term for the classical Riemann zeta function  has an extensive literature. See, for example \cite{G}, \cite{Fu} and the bibliography there. Here are some pictures to convey a sense of how our $S_E(X;T)$ behaves, at least in the currently computable range which (roughly) allows $T$ to be only as high as  $10^4$.
\subsection{The bias between under-counts and over-counts}
  We will assume that our data has `Explicit Formula' statistics, and---copying Sarnak ({\cite{S}})--- apply this to the question we began with, i.e., what is the ``bias" in the race between under-counts and over-counts?

@@

[[idea -- instead just define $\Phi(H,T)$ to be the min
of the finitely many possible linear combinations, and ask
about the behavior of $\Phi$.  Is there some clever algorithm
to get information about this?]]
\end{itemize}\end{itemize}
\end{document}
  \centerline{ $E = 11a$}
 \vskip10pt
 \includegraphics[width=0.9\textwidth]{plots/OSC11.pdf}

 %More specifically, the {\it means}  directly interpretable as {\it biases}  determine   the orders of vanishing at the central point of the $L$-functions of the symmetric powers of odd order of the modular eigenform attached to $E$ and vice versa..
